Building Markov State Model for a Periodically Driven Non-Equilibrium System

### 2014-07-14 writedown/period.noneq.msm/main.tex

\documentclass[aip,jcp,preprint,unsortedaddress,a4paper,onecolumn]{revtex4-1}
% \documentclass[acs, jctcce, a4paper,preprint,unsortedaddress,onecolumn]{revtex4-1}
% \documentclass[aps,pre,twocolumn,unsortedaddress]{revtex4-1}
% \documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

@@

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{red} #1}}
\newcommand{\greenc}[1]{{\color{green} #1}}
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dd}[1]{\textsf{#1}}
\newcommand{\fwd}[0]{\textrm{fwd}}
\newcommand{\bwd}[0]{\textrm{bwd}}
\newcommand{\period}[0]{T_{\textrm{P}}}
\newcommand{\ml}[0]{\mathcal {L}}
\newcommand{\mo}[0]{\mathcal {O}}
\newcommand{\mbp}[0]{\mathbb {P}}
\newcommand{\dt}[0]{\Delta t}

@@

Now we assume the overdamped dynamics
\begin{align}
  dx_t = -\nabla_x V(x_t,t) dt + \sigma dw_t
\end{align}
where $V$ is the time-dependent and $t$-periodic potential. For simplicity we denote the force by $F(x_t,t) = -\nabla_x V(x_t, t)$. $dw_t$ is
the standard Wiener process.  Assuming a discretization of the
stochastic process at time $0 < t_1 < t_2 < \cdots < t_N = T$, where
$t_i = iT / N$. We denote $x_i = x_{t_i}$, and $w_i = w_{t_i}$, then we have the
conditional probability:
\begin{align}\nonumber
  p(x_T,T\vert x_0,0) =\lim_{N\rightarrow\infty} &\,
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_1\cdots\int dx_{N-1}\\\nonumber
  &\,
  \exp\bigg\{\frac1\sigma\sum_{i=0}^{N-1} F(x_i,t_i)(w_{i+1} - w_{i}) - \frac1{2\sigma^2}\sum_{i=0}^{N-1}F^2(x_i,t_i)\dt\bigg\} \times\\\label{eqn:tmp8}
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x_{i+1} - x_i)^2}{\dt}\bigg\}.
\end{align}
Now consider the differentiation:
\begin{align}\nonumber
  F(x_i,t_i) - F(x_{i+1},t_{i+1}) =
  &\,
  F(x_i,t_i) - F(x_{i+1},t_{i}) + F(x_{i+1},t_{i}) -  F(x_{i+1},t_{i+1})\\\nonumber
  =&\,
  -\sigma\nabla_x F(x_{i+1},t_{i})dw_t + \mo(\dt) \\\label{eqn:tmp9}
  =&\,
  -\sigma\nabla_x F(x_{i+1},t_{i+1})dw_t + \mo(\dt) 
\end{align}
and
\begin{align}\label{eqn:tmp10}
  \sigma(w_{i+1} - w_{i}) =
  &\,
  x_{i+1} - x_i 
\end{align}
Inserting Eq.~\eqref{eqn:tmp9} and \eqref{eqn:tmp10} into Eq.~\eqref{eqn:tmp8}, all terms of order $\mo(\dt)$ vanish at the limit $N\rightarrow\infty$,
% so with the formal identity $(dw_t)^2 = dt$,  we have
\begin{align}\nonumber
  p(x_T,T\vert x_0,0) =\lim_{N\rightarrow\infty} &\,
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_1\cdots\int dx_{N-1}\\\nonumber
  &\,
  \exp\bigg\{\frac1{\sigma^2}\sum_{i=0}^{N-1} F(x_{i+1},t_{i+1})(x_{i+1} - x_i) \bigg\} \times \\ \nonumber
  &\,
  \exp\bigg\{-\frac{1}{\sigma^2}\sum_{i=0}^{N-1}  \nabla_x F(x_{i+1},t_{i+1}) (x_{i+1} - x_{i})^2\bigg\} \times \\\nonumber
  &\,
  \exp\bigg\{-\frac1{2\sigma^2}\sum_{i=0}^{N-1}F^2(x_i,t_i)\dt\bigg\} \times\\\label{eqn:tmp11}
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x_{i+1} - x_i)^2}{\dt}\bigg\}.
\end{align}
Now, consider a trajectory that starts at $x_T$, ends at $x_0$, and visits $x_{N-i}$ at time $t_i$, the conditional probability
gives
\begin{align}\nonumber
  p(x_0,T\vert x_T,0)
  =\lim_{N\rightarrow\infty} &\,
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_{N-1}\cdots\int dx_{1}\\\nonumber
  &\,
  \exp\bigg\{\frac1\sigma\sum_{i=N}^{1} F(x_i,t_{N-i})(x_{i-1} - x_{i}) - \frac1{2\sigma^2}\sum_{i=N}^{1}F^2(x_i,t_{N-i})\dt\bigg\} \times\\\nonumber
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=N}^{1} \frac{(x_{i-1} - x_{i})^2}{\dt}\bigg\} \\\nonumber
  =\lim_{N\rightarrow\infty} &\,
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_{1}\cdots\int dx_{N}\\\nonumber
  &\,
  \exp\bigg\{\frac1\sigma\sum_{i=0}^{N-1} -F(x_{i+1},t_{i+1})(x_{i} - x_{i+1}) - \frac1{2\sigma^2}\sum_{i=0}^{N-1}F^2(x_{i+1},t_{i+1})\dt\bigg\} \times\\\nonumber
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x_{i} - x_{i+1})^2}{\dt}\bigg\}\\\nonumber
  =\lim_{N\rightarrow\infty} &\,
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_{1}\cdots\int dx_{N}\\\nonumber
  &\,
  \exp\bigg\{\frac1\sigma\sum_{i=0}^{N-1} F(x_{i+1},t_{i+1})(x_{i+1} - x_{i})\bigg\}\times\\\nonumber
  &\,
  \exp\bigg\{- \frac1{2\sigma^2}\sum_{i=0}^{N-1}F^2(x_{i+1},t_{i+1})\dt\bigg\} \times\\\nonumber
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x_{i} - x_{i+1})^2}{\dt}\bigg\}.
\end{align}

% Now we consider the time-reversed process $y_{N-i} = x_i$,
% \begin{align}\nonumber
%   p(y_0,T\vert y_N,0) =\lim_{N\rightarrow\infty} &\,
%   \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dy_1\cdots\int dx_{y-1}\\\nonumber
%   &\,
%   \exp\bigg\{\frac1\sigma\sum_{i=0}^{N-1} F(y_{N - i-1},t_{N-i-1})(w_{N-i-1} - w_{N-i}) \bigg\} \times \\ \nonumber
%   &\,
%   \exp\bigg\{-\sum_{i=0}^{N-1}  \nabla_x F(x_{i+1},t_{i+1})\dt\bigg\} \times \\\nonumber
%   &\,
%   \exp\bigg\{-\frac1{2\sigma^2}\sum_{i=0}^{N-1}F^2(x_i,t_i)\dt\bigg\} \times\\\label{eqn:tmp8}
%   &\,
%   \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x_{i+1} - x_i)^2}{\dt}\bigg\}.
% \end{align}

We notice that
\begin{align}\nonumber
  dV(x, t) = &\, \frac{\partial V}{\partial x} dx + \frac{\partial V}{\partial t} dt\\\nonumber
  =&\,
  (\nabla_x V F + \frac12 \sigma^2 \nabla^2_x V) dt + \sigma \nabla V dw_t + \frac{\partial V}{\partial t} dt \\\nonumber
  =&\,
  -F^2dt + \frac12 \sigma^2\nabla_x F dt - \sigma F dw_t + \frac{\partial V}{\partial t} dt 
\end{align}

>>>>

\documentclass[preprint,unsortedaddress,a4paper,onecolumn]{revtex4-1}
% \documentclass[acs, jctcce, a4paper,preprint,unsortedaddress,onecolumn]{revtex4-1}
% \documentclass[aps,pre,twocolumn,unsortedaddress]{revtex4-1}
% \documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

@@

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{red} #1}}
\newcommand{\greenc}[1]{{\color{green} #1}}
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dd}[1]{\textsf{#1}}
\newcommand{\fwd}[0]{\textrm{fwd}}
\newcommand{\bwd}[0]{\textrm{bwd}}
\newcommand{\period}[0]{T_{\textrm{P}}}
\newcommand{\ml}[0]{\mathcal {L}}
\newcommand{\mo}[0]{\mathcal {O}}
\newcommand{\mbp}[0]{\mathbb {P}}
\newcommand{\mc}[0]{\mathcal {C}}
\newcommand{\dt}[0]{\Delta t}

@@

Now we assume the overdamped dynamics
\begin{align}\label{eqn:tmp7}
  dx_t = -\nabla_x V(x_t,t) dt + \sigma dw_t
\end{align}
where $V$ is the time-dependent and $t$-periodic potential. For simplicity we denote the force by $F(x_t,t) = -\nabla_x V(x_t, t)$. $dw_t$ is
the standard Wiener process.  Assuming a discretization of the
stochastic process at time $0 < t_1 < t_2 < \cdots < t_N = T$, where
$t_i = iT / N$. We denote $x_i = x_{t_i}$, and $w_i = w_{t_i}$, then we have the
conditional probability:
\begin{align}\nonumber
  p(x_T,T\vert x_0,0) =\lim_{N\rightarrow\infty} &\,
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_1\cdots\int dx_{N-1}\\\nonumber
  &\,
  \exp\bigg\{\frac1\sigma\sum_{i=0}^{N-1} F(x_i,t_i)(w_{i+1} - w_{i}) - \frac1{2\sigma^2}\sum_{i=0}^{N-1}F^2(x_i,t_i)\dt\bigg\} \times\\\label{eqn:tmp8}
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x_{i+1} - x_i)^2}{\dt}\bigg\}.
\end{align}
Notice that
\begin{align}\label{eqn:tmp10}
  \sigma(w_{i+1} - w_{i}) =
  &\,
  x_{i+1} - x_i 
\end{align}
we have
% so with the formal identity $(dw_t)^2 = dt$,  we have
\begin{align}\nonumber
  p(x_T,T\vert x_0,0)
  =\lim_{N\rightarrow\infty} &\,
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_1\cdots\int dx_{N-1}\\\nonumber
  &\,
  \exp\bigg\{\frac1{\sigma^2}\sum_{i=0}^{N-1} F(x_{i},t_{i})(x_{i+1} - x_i) -\frac1{2\sigma^2}\sum_{i=0}^{N-1}F^2(x_i,t_i)\dt\bigg\} \times\\\label{eqn:tmp10}
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x_{i+1} - x_i)^2}{\dt}\bigg\}.
\end{align}
Now, consider a conjugate trajectory $x^\dagger_t = x_{T-t}$ that starts at $x_T$, ends at $x_0$. The conjugate dynamics is driven by  $F^\dagger(x^\dagger_t,t) = F(x^\dagger_t, T-t)$.
Since we have the symmetry $F(x_t, -t) = -F(x_t, t)$, the conjugate dynamics also satisfies Eq.~\eqref{eqn:tmp7}.
Therefore, we have the probability:
\begin{align}\nonumber
  p(x_0,T\vert x_T,0) = &\,
  p^\dagger(x^\dagger_T,T\vert x^\dagger_0,0)  \\\nonumber
  = &\,
  \lim_{N\rightarrow\infty}
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx^\dagger_1\cdots\int dx^\dagger_{N-1}\\\nonumber
  &\,
  \exp\bigg\{\frac1{\sigma^2}\sum_{i=0}^{N-1} F^\dagger(x^\dagger_{i},t_{i})(x^\dagger_{i+1} - x^\dagger_i) -\frac1{2\sigma^2}\sum_{i=0}^{N-1}[F^\dagger(x^\dagger_i,t_i)]^2\dt\bigg\} \times\\\nonumber
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=0}^{N-1} \frac{(x^\dagger_{i+1} - x^\dagger_i)^2}{\dt}\bigg\} \\\nonumber
  =&\,
  \lim_{N\rightarrow\infty} 
  \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_1\cdots\int dx_{N-1}\\\nonumber
  &\,
  \exp\bigg\{\frac1{\sigma^2}\sum_{i=N}^{1} F(x_{i},t_{i})(x_{i-1} - x_i) -\frac1{2\sigma^2}\sum_{i=N}^{1}F^2(x_i,t_i)\dt\bigg\} \times\\\label{eqn:tmp11}
  &\,
  \exp\bigg\{- \frac1{2\sigma^2} \sum_{i=N}^{1} \frac{(x_{i-1} - x_i)^2}{\dt}\bigg\}.
\end{align}
The difference between the single trajectory probabilities is
\begin{align}\label{eqn:tmp12}
  \log\bigg\{\frac{  p[x_t] }{ p[x^\dagger_t]} \bigg\}
  =&\,
  \frac1{\sigma^2}\sum_{i=1}^{N-1} F(x_i,t_i)(x_{i+1} - x_{i}) + F(x_i,t_i)(x_{i} - x_{i-1}) 
\end{align}
Now consider the differentiation:
\begin{align}\nonumber
  F(x_{i},t_{i}) - F(x_{i-1},t_{i-1}) =
  &\,
  F(x_{i},t_{i}) - F(x_{i-1},t_{i}) + F(x_{i-1},t_{i}) -  F(x_{i-1},t_{i-1})\\\nonumber
  =&\,
  \nabla_x F(x_{i-1},t_{i})(x_i - x_{i-1}) + \mo(\dt) \\\label{eqn:tmp13}
  =&\,
  \nabla_x F(x_{i-1},t_{i-1})(x_i - x_{i-1}) + \mo(\dt)
\end{align}
Then the \eqref{eqn:tmp12} becomes
\begin{align}\label{eqn:tmp14}
  \log\bigg\{\frac{  p[x_t] }{ p[x^\dagger_t]} \bigg\}
  =&\,
  \frac2{\sigma^2}\sum_{i=0}^{N-1} F(x_i,t_i)(x_{i+1} - x_{i}) +
  \frac1{\sigma^2}\sum_{i=0}^{N-1}\nabla_xF(x_i,t_i)(x_{i+1} - x_{i})^2
\end{align}
We notice that
\begin{align}\nonumber
  dV(x, t) = &\, \frac{\partial V}{\partial x} dx + \frac{\partial V}{\partial t} dt\\\nonumber
  =&\,
  \frac12 \sigma^2 \nabla^2_x V dt +  \nabla V dx_t + \frac{\partial V}{\partial t} dt \\
  =&\,
  -\frac12 \sigma^2 \nabla_x F dt -  F dx_t + \frac{\partial V}{\partial t} dt
\end{align}
and the identity
\begin{align}
  dt = (dw_t)^2 = \frac1{\sigma^2} dx_t^2
\end{align}
Eq.~\eqref{eqn:tmp14} becomes
\begin{align}
  \log\bigg\{\frac{  p[x_t] }{ p[x^\dagger_t]} \bigg\}
  =&\,
  -\frac2{\sigma^2}\bigg[
  V(x_N,t_N) - V(x_0,t_0) + \sum_{i=0}^{N-1}\partial_tV(x_i,t_i)\dt
  \bigg]
\end{align}
According to the  Einstein relation $T = \sigma^2/2$, by denoting $\beta = 1/T$
\begin{align}
  \log\bigg\{\frac{  p[x_t] }{ p[x^\dagger_t]} \bigg\}
  =&\,
  -\beta\bigg[
  V(x_N,t_N) - V(x_0,t_0) + \sum_{i=0}^{N-1}\partial_tV(x_i,t_i)\dt
  \bigg]  
\end{align}
Therefore
\begin{align}\nonumber
  p(x_0,T\vert x_T,0)
  =&\,
  % \lim_{N\rightarrow\infty} 
  % \frac{1}{(2\pi\sigma^2\dt)^{(N-1)/2}} \int dx_{N-1}\cdots\int dx_{1}
  \int_{\mc\{x_0,0;x_T,T\}}
  \frac{  p[x^\dagger_t] }{ p[x_t]} \cdot p[x_t] \\\label{eqn:tmp19}
  = &\,
  \frac{e^{-\beta V(x_0,t_0)}}{e^{-\beta V(x_N,t_N)}}
  \int_{\mc\{x_0,0;x_T,T\}}
  \exp\bigg\{
  \beta\int_0^T \partial_t V(x_t,t)dt 
  \bigg\} \cdot p[x_t]
\end{align}
where $\mc\{x_0,0;x_T,T\}$ denotes all continuous trajectories starting at $x_0$ and ending at $x_T$.
If $\partial_t V = 0$, i.e.~equilibrium, we have
\begin{align}
  p(x_0,T\vert x_T,0)e^{-\beta V(x_T,T)} =  p(x_T,T\vert x_0,0) e^{-\beta V(x_0,0)},
\end{align}
which proves the reversibility of the equilibrium dynamics.
The term
\begin{align}
  W[x_t] = - \int_0^T \partial_t V(x_t,t)dt
\end{align}
is the non-equilibrium work associated to all possible
the dynamics $x_t$ starting at $x_0$ and ending at $x_T$ (see e.g.~Ref.~\cite{seifert2012stochastic}).
Therefore Eq.~\eqref{eqn:tmp19} is the detailed Jarzynski relation.
Now the problem becomes if we can write a nice form (e.g.~the difference of a state function measured at $x_T$ and $x_0$) for the non-equilibrium work.
